Periodograms for Multiband Astronomical Time Series

### 2015-02-04 writeup/paper.tex
paper: spell check and add small changes

\begin{abstract}
This paper introduces the {\it multiband periodogram}, a general extension of the well-known 
Lomb-Scargle approach for detecting periodic signals in time-domain data. In addition to 
advantanges of the Lomb-Scargle method such as treatment of non-uniform sampling and
heteroscedastic errors, the multiband periodogram significantly improves period finding for randomly 
sampled multiband light curves (e.g., Pan-STARRS, DES and LSST). The light curves in 
each band are modeled as arbitrary truncated Fourier series, with the period and phase 
shared across all bands. The key aspect is the use of Tikhonov regularization which drives
most of the variability into the so-called base model common to all bands, while 
fits for individual bands describe residuals relative to the base model and typically require
lower-order Fourier series. This decrease in the required number of fit parameters is the
main reason for improved performance. We use simulated light curves 
and randomly subsampled SDSS Stripe 82 data to demonstrate the superiority of this 
method compared to other methods from the literature, and find that this method will be able to
efficiently determine the correct period in the majority of LSST's bright RR Lyrae stars with
as little as six months of LSST data.
A Python implementation of this method, along with code to fully reproduce the results
reported here, is available on GitHub.
\end{abstract}

@@

The most popular parametric method is the Lomb-Scargle periodogram, which is discussed in detail in \sect{brief_overview}.
The Lomb-Scargle periodogram is related to the $\chi^2$ for a least-square fit of a single sinusoid to data
and can treat non-uniformly sampled time series with heteroscedastic measurement uncertainties. 
The underlying model of the Lomb–Scargle periodogram is nonlinear in frequency and basis functions at different
frequencies are not orthogonal. As a result, the periodogram has many local maxima and thus in practice the global 
maximum of the periodogram is found by grid search \citep[for details see, e.g.][]{ICVG2014}.
A more general parameteric method based on the use of continuous-time autoregressive moving average (CARMA) model
was recently introduced by \citet{Kelly14}. CARMA models can also treat non-uniformly sampled time series with 
heteroscedastic measurement uncertainties, and can handle complex variability patterns.

A weakness of the above methods is that they require homogeneous measurements -- for astronomy data, this means 
that successive measurements must be taken through a single photometric bandpass (filter). This has not been a major
problem for past surveys because measurements are generally taken through a single photometric filter 
\citep [e.g. LINEAR,][]{LINEAR1}, or nearly-simultaneously in all bands at each observation \citep [e.g. SDSS,][]{Sesar2010}.
For the case of simultaneously taken multiband measurements, \cite{Suveges12} utilized the principal component
method to optimaly extract the best period. Their method is essentially a multiband generalization of the well-known
two-band Welch-Stetson variability index \citep{Stetson1996}. Unfortunately, when data in each band are taken at
different times, the  principal component approach in not applicable. In such cases, past studies have generally relied 
on \adhoc{} methods such as a majority vote among multiple single-band estimates of the 
periodogram \citep[e.g.,][]{Oluseyi12}.

@@

Because astronomical observing cadences are rarely so uniform, many have looked at extending the ideas behind the periodogram to work with unevenly-sampled data. Most famously, \citet{Lomb76} and \citet{Scargle82} extended earlier work to define the {\it normalized periodogram}:
\begin{equation}
  \eqlabel{LombScargle}
  P_N(\omega) = \frac{1}{2\,V_y}\left[
    \frac{\left[\sum_k(y_k - \bar{y})\cos\omega(t_k - \tau)\right]^2}
    {\sum_k \cos^2\omega(t_k - \tau)}
    +
    \frac{\left[\sum_k(y_k - \bar{y})\sin\omega(t_k - \tau)\right]^2}
    {\sum_k \sin^2\omega(t_k - \tau)}
\right],
\end{equation}
where $\bar{y}$ is the mean and $V_y$ is the variance of the data $\{y_k\}$, and $\tau$ is the time-offset which makes $P_N(\omega)$ independent of a translation in $t$ \citep[see][for an in-depth discussion]{NumRec}. \citet{Lomb76} showed that this time-offset has a deeper effect: namely, it makes $P_N$ identical to the estimate of harmonic content given a least-squares fit to a single-component sinusoidal model,
\begin{equation}
  \eqlabel{SingleModel}
  d(t) = A\sin(\omega t + \phi).
\end{equation}
This long-recognized connection between spectral power and least squares fitting methods was solidified by \citet{Jaynes87}, who demonstrated that the normalized periodogram of Lomb and Scargle is a sufficient statistic for inferences about a stationary-frequency signal in the presence of Gaussian noise. Building on this result, \citet{Bretthorst88} explored the extension of these methods to more complicated models with multiple frequency terms, nonstationary frequencies, and other more sophisticated models within a Bayesian framework.

@@

\begin{eqnarray}
  \chi^2(\omega) &=& (y - X_\omega\theta)^T\Sigma^{-1}(y - X_\omega\theta)\\
  \chi^2_0 &=& y^T \Sigma^{-1} y
\end{eqnarray}
The expression for the normalized periodogram can be computed by finding via standard methods the value of $\theta$ which minizes $\chi^2(\omega)$, and plugging the result into \eq{chi2PN}. This yields
\begin{equation}
  \eqlabel{LombScargle2}
  P_N(\omega) = \frac{y^T\Sigma^{-1}X_\omega~[X_\omega^T\Sigma^{-1}X_\omega]^{-1}~X_\omega^T\Sigma^{-1}y}{y^T\Sigma^{-1}y}.
\end{equation}
We note that this expression is equivalent to \eq{LombScargle} in the homoscedastic case with $\Sigma \propto V_y I$.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig02.pdf}
  \caption{
    An illustration of the effect of the floating mean model for censored data.
    The data consist of 80 observations drawn from a sinusoidal model. To mimick a potentially damaging selection effect, all observations with magnitude fainter than 16 are removed (indicated by the light-gray points). The standard and floating-mean periodograms are computed from the remaining data; these fits are shown over the data in the left panel. Because of this biased observing pattern, the mean of the observed data is a biased estimator of the true mean. The standard fixed-mean model in this case fails to recover the true period of 0.622 days, while the floating mean model still finds the correct period.
  }
  \figlabel{floating_mean}
\end{figure}

As an example of one of these generalizations, we'll consider what \citet{Zechmeister09} call the ``generalized Lomb-Scargle method'', and which we'll call the {\it floating-mean periodogram}. This method adjusts the classic normalized periodogram by fitting the mean of the model alongside the amplitudes:
\begin{equation}
  y(t~|~\omega, \theta) = \theta_0 + \theta_1\sin\omega t + \theta_2\cos\omega t
\end{equation}
This model can be more accurate than the standard periodogram for certain observing cadences and selection functions. \citet{Zechmeister09} detail the required modifications to the harmonic formalism of \eq{LombScargle} to allow the mean to float in the model. In the matrix formalism, the modification is much more straightforward: all that is required is to add a column of ones to the $X_\omega$ matrix before computing the power via \eq{LombScargle2}. This column of ones corresponds to a third entry in the parameter vector $\theta$, and acts as a uniform constant offset for all datapoints.

@@

The lower panels of \fig{multiband_sim} show the same 60 nights of data, except with only a {\it single} band observation recorded each night.
The lower-left panel shows the observations as a function of phase, and the lower-right panels show the periodograms derived from the data.
With only 12 observations for each indiviudal band, it is clear that there is not enough data to accurately determine the period within each single band. The shared-phase multiband approach (i.e. $N_{base}=1,N_{band}=0$), shown in the lower-right panel, fits a single model to the full data and clearly recovers the true frequency of 0.622 days.

@@

\section{Application to Stripe 82 RR Lyrae}
\sectlabel{stripe82}
Stripe 82 is a three hundred square degree equatorial region of the sky which was repeatedly imaged through multiple band-passes during phase II of the Sloan Digital Sky Survey \citep[SDSS II, see][]{Sesar2007}.
Here we consider the SDSS II observations of 483 RR Lyrae stars compiled and studied by S10, in which periods for these stars were determined based on emperically-derived light curve templates.
Because the template-fitting method is extremely computationally intensive, S10 first determined candidate periods by taking the top 5 results of the Supersmoother \citep{Reimann94} algorithm applied to the $g$-band; template fits were then performed at each candidate period and the period with the best template fit was reported as the true period. In this section, we make use of this dataset to quantitatively evaluate the effectiveness of the multiband periodogram approach.

@@

These results point to much more promising prospects for science with variable stars than previously reported.
In particular, even with only six months of LSST data, we can expect to correctly identify the periods for over 60\% of stars brighter than $g=22$; with the first two years of LSST observations, this increases to nearly 100\%.
Part of this improvement is due to the performance of the shared-phase multiband model with noisy data, and part of this improvement is due to the relaxed period-matching constraints enabled by the hybrid approach of periodogram-based and template-based period determination.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{fig09.pdf}
  \caption{
    Fraction of periods correctly determined for LSST RR Lyrae as a function
    of the length of the observing season and the mean $g$-band magnitude, for the multiband (solid lines)
    and single-band supersmoother (dashed lines) approaches.
  } 
  \figlabel{LSST_sims}
\end{figure}

@@

{\it Acknowledgements:} JTV is supported by the University of Washington eScience institute, including grants from the Alfred P. Sloan Foundation, the Gordon and Betty Moore Foundation, and the Washington Research Foundation. The authors thank GitHub for providing free academic accounts which were essential in the development of this work.

@@

\appendix
\section{Python Implementation of Multiband Periodogram}
\sectlabel{gatspy}
The algorithm outlined in this paper is available in {\tt gatspy}, an open-source Python package for general astronomical time-series analysis\footnote{\url{http://github.com/jakevdp/gatspy/}}. Along with the periodogram implementation, it also contains code to download all the data used in this work. Code to reproduce this paper, including all figures, is available in a separate repository\footnote{\url{http://github.com/jakevdp/multiband_LS/}}.

@@

The models in the \texttt{gatspy} package contain many more methods, and much more functionality that what is shown here. For updates, more examples, and more information, visit \url{http://github.com/jakevdp/gatspy/}.

>>>>

\begin{abstract}
This paper introduces the {\it multiband periodogram}, a general extension of the well-known 
Lomb-Scargle approach for detecting periodic signals in time-domain data. In addition to 
advantages of the Lomb-Scargle method such as treatment of non-uniform sampling and
heteroscedastic errors, the multiband periodogram significantly improves period finding for randomly 
sampled multiband light curves (e.g., Pan-STARRS, DES and LSST). The light curves in 
each band are modeled as arbitrary truncated Fourier series, with the period and phase 
shared across all bands. The key aspect is the use of Tikhonov regularization which drives
most of the variability into the so-called base model common to all bands, while 
fits for individual bands describe residuals relative to the base model and typically require
lower-order Fourier series. This decrease in the required number of fit parameters is the
main reason for improved performance. We use simulated light curves 
and randomly subsampled SDSS Stripe 82 data to demonstrate the superiority of this 
method compared to other methods from the literature, and find that this method will be able to
efficiently determine the correct period in the majority of LSST's bright RR Lyrae stars with
as little as six months of LSST data.
A Python implementation of this method, along with code to fully reproduce the results
reported here, is available on GitHub.
\end{abstract}

@@

The most popular parametric method is the Lomb-Scargle periodogram, which is discussed in detail in \sect{brief_overview}.
The Lomb-Scargle periodogram is related to the $\chi^2$ for a least-square fit of a single sinusoid to data
and can treat non-uniformly sampled time series with heteroscedastic measurement uncertainties. 
The underlying model of the Lomb–Scargle periodogram is nonlinear in frequency and basis functions at different
frequencies are not orthogonal. As a result, the periodogram has many local maxima and thus in practice the global 
maximum of the periodogram is found by grid search \citep[for details see, e.g.][]{ICVG2014}.
A more general parametric method based on the use of continuous-time autoregressive moving average (CARMA) model
was recently introduced by \citet{Kelly14}. CARMA models can also treat non-uniformly sampled time series with 
heteroscedastic measurement uncertainties, and can handle complex variability patterns.

A weakness of the above methods is that they require homogeneous measurements -- for astronomy data, this means 
that successive measurements must be taken through a single photometric bandpass (filter). This has not been a major
problem for past surveys because measurements are generally taken through a single photometric filter 
\citep [e.g. LINEAR,][]{LINEAR1}, or nearly-simultaneously in all bands at each observation \citep [e.g. SDSS,][]{Sesar2010}.
For the case of simultaneously taken multiband measurements, \cite{Suveges12} utilized the principal component
method to optimally extract the best period. Their method is essentially a multiband generalization of the well-known
two-band Welch-Stetson variability index \citep{Stetson1996}. Unfortunately, when data in each band are taken at
different times, the  principal component approach in not applicable. In such cases, past studies have generally relied 
on \adhoc{} methods such as a majority vote among multiple single-band estimates of the 
periodogram \citep[e.g.,][]{Oluseyi12}.

@@

Because astronomical observing cadences are rarely so uniform, many have looked at extending the ideas behind the periodogram to work with unevenly-sampled data. Most famously, \citet{Lomb76} and \citet{Scargle82} extended earlier work to define the {\it normalized periodogram}:
\begin{equation}
  \eqlabel{LombScargle}
  P_N(\omega) = \frac{1}{2\,V_y}\left[
    \frac{\left[\sum_k(y_k - \bar{y})\cos\omega(t_k - \tau)\right]^2}
    {\sum_k \cos^2\omega(t_k - \tau)}
    +
    \frac{\left[\sum_k(y_k - \bar{y})\sin\omega(t_k - \tau)\right]^2}
    {\sum_k \sin^2\omega(t_k - \tau)}
\right],
\end{equation}
where $\bar{y}$ is the mean and $V_y$ is the variance of the data $\{y_k\}$, and $\tau$ is the time-offset which makes $P_N(\omega)$ independent of a translation in $t$ \citep[see][for an in-depth discussion]{NumRec}. \citet{Lomb76} showed that this time-offset has a deeper effect: namely, it makes $P_N$ identical to the estimate of harmonic content given a least-squares fit to a single-component sinusoidal model,
\begin{equation}
  \eqlabel{SingleModel}
  d(t) = A\sin(\omega t + \phi).
\end{equation}
This long-recognized connection between spectral power and least squares fitting methods was solidified by \citet{Jaynes87}, who demonstrated that the normalized periodogram of Lomb and Scargle is a sufficient statistic for inferences about a stationary-frequency signal in the presence of Gaussian noise. Building on this result, \citet{Bretthorst88} explored the extension of these methods to more complicated models with multiple frequency terms, non-stationary frequencies, and other more sophisticated models within a Bayesian framework.

@@

\begin{eqnarray}
  \chi^2(\omega) &=& (y - X_\omega\theta)^T\Sigma^{-1}(y - X_\omega\theta)\\
  \chi^2_0 &=& y^T \Sigma^{-1} y
\end{eqnarray}
The expression for the normalized periodogram can be computed by finding via standard methods the value of $\theta$ which minimizes $\chi^2(\omega)$, and plugging the result into \eq{chi2PN}. This yields
\begin{equation}
  \eqlabel{LombScargle2}
  P_N(\omega) = \frac{y^T\Sigma^{-1}X_\omega~[X_\omega^T\Sigma^{-1}X_\omega]^{-1}~X_\omega^T\Sigma^{-1}y}{y^T\Sigma^{-1}y}.
\end{equation}
We note that this expression is equivalent to \eq{LombScargle} in the homoscedastic case with $\Sigma \propto V_y I$.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig02.pdf}
  \caption{
    An illustration of the effect of the floating mean model for censored data.
    The data consist of 80 observations drawn from a sinusoidal model. To mimic a potentially damaging selection effect, all observations with magnitude fainter than 16 are removed (indicated by the light-gray points). The standard and floating-mean periodograms are computed from the remaining data; these fits are shown over the data in the left panel. Because of this biased observing pattern, the mean of the observed data is a biased estimator of the true mean. The standard fixed-mean model in this case fails to recover the true period of 0.622 days, while the floating mean model still finds the correct period.
  }
  \figlabel{floating_mean}
\end{figure}

As an example of one of these generalizations, we'll consider what \citet{Zechmeister09} call the ``generalized Lomb-Scargle method'', and which we'll call the {\it floating-mean periodogram}. This method adjusts the classic normalized periodogram by fitting the mean of the model alongside the amplitudes:
\begin{equation}
  y(t~|~\omega, \theta) = \theta_0 + \theta_1\sin\omega t + \theta_2\cos\omega t
\end{equation}
This model can be more accurate than the standard periodogram for certain observing cadences and selection functions. \citet{Zechmeister09} detail the required modifications to the harmonic formalism of \eq{LombScargle} to allow the mean to float in the model. In the matrix formalism, the modification is much more straightforward: all that is required is to add a column of ones to the $X_\omega$ matrix before computing the power via \eq{LombScargle2}. This column of ones corresponds to a third entry in the parameter vector $\theta$, and acts as a uniform constant offset for all data points.

@@

The lower panels of \fig{multiband_sim} show the same 60 nights of data, except with only a {\it single} band observation recorded each night.
The lower-left panel shows the observations as a function of phase, and the lower-right panels show the periodograms derived from the data.
With only 12 observations for each individual band, it is clear that there is not enough data to accurately determine the period within each single band. The shared-phase multiband approach (i.e. $N_{base}=1,N_{band}=0$), shown in the lower-right panel, fits a single model to the full data and clearly recovers the true frequency of 0.622 days.

@@

\section{Application to Stripe 82 RR Lyrae}
\sectlabel{stripe82}
Stripe 82 is a three hundred square degree equatorial region of the sky which was repeatedly imaged through multiple band-passes during phase II of the Sloan Digital Sky Survey \citep[SDSS II, see][]{Sesar2007}.
Here we consider the SDSS II observations of 483 RR Lyrae stars compiled and studied by S10, in which periods for these stars were determined based on empirically-derived light curve templates.
Because the template-fitting method is extremely computationally intensive, S10 first determined candidate periods by taking the top 5 results of the Supersmoother \citep{Reimann94} algorithm applied to the $g$-band; template fits were then performed at each candidate period and the period with the best template fit was reported as the true period. In this section, we make use of this dataset to quantitatively evaluate the effectiveness of the multiband periodogram approach.

@@

These results point to much more promising prospects for science with variable stars than previously reported.
In particular, even with only six months of LSST data, we can expect to correctly identify the periods for over 60\% of stars brighter than $g=22$; with the first two years of LSST observations, this increases to nearly 100\%; with five years of data, the multiband method identifies the correct period for 100\% of even the faintest stars.
Part of this improvement is due to the performance of the shared-phase multiband model with noisy data, and part of this improvement is due to the relaxed period-matching constraints enabled by the hybrid approach of periodogram-based and template-based period determination.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{fig09.pdf}
  \caption{
    Fraction of periods correctly determined for LSST RR Lyrae as a function
    of the length of the observing season and the mean $g$-band magnitude, for the multiband (solid lines)
    and single-band supersmoother (dashed lines) approaches.
    The multiband method is superior to the single-band supersmoother approach in all cases, and especially for the faintest objects.
  } 
  \figlabel{LSST_sims}
\end{figure}

@@

{\it Acknowledgments:} JTV is supported by the University of Washington eScience institute, including grants from the Alfred P. Sloan Foundation, the Gordon and Betty Moore Foundation, and the Washington Research Foundation. The authors thank GitHub for providing free academic accounts which were essential in the development of this work.

@@

\appendix
\section{Python Implementation of Multiband Periodogram}
\sectlabel{gatspy}
The algorithm outlined in this paper is available in {\tt gatspy}, an open-source Python package for general astronomical time-series analysis\footnote{\url{http://github.com/astroml/gatspy/}} \citep{gatspy}. Along with the periodogram implementation, it also contains code to download all the data used in this work. Code to reproduce this paper, including all figures, is available in a separate repository\footnote{\url{http://github.com/jakevdp/multiband_LS/}}.

@@

The models in the \texttt{gatspy} package contain many more methods, and much more functionality that what is shown here. For updates, more examples, and more information, visit \url{http://github.com/astroml/gatspy/}.
